Liveness-Based Garbage Collection

### 2014-09-10 paper/sections/syntax-and-semantics.tex
Op sematics first version

\subsubsection{Semantics}
We now give an operational
semantics for  our language.  Later, a refinement of
the  operational semantics,  which we  call minefield  semantics, will
serve to prove liveness analysis correct.
We give a small-step semantics because, unlike big-step semantics,
correctness for non-terminating programs does not need special treatment.
We start with the domains used by the semantics:
\[
\begin{array}{rlcl@{\hspace{2em}}l}
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Loc} & \mbox{-- Environment} \\ 
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Val \times
  Val}& \mbox{-- Values}\\
c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (\mathit{Val} +
\mathit{Clo}) & \mbox{-- Heap}
\end{array}
\]
%Note that in the eager case, if a cons had a numeric argument, this
%fact was known when the cons was created. Thus you could have unboxed
%numbers. This is not possible now. For instance, the argument of a
%cons could be a closure which evaluates to a number.
%Decision - numbers are boxed
Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
A value is either a number, the empty list $\NIL$, or a location  $\ell$.
Our liveness analysis does not track numeric values, and thus is neutral
as to whether these are boxed or represented as immediates.
An environment is a finite mapping from variables to values, and a heap a finite mapping
from locations to pairs of values.
% Given  a program,  an  environment is  a
% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
% values in $\mathit{Val}$.    The heap is a mapping
% from  locations  to pairs of  values.
Finally, $\stk$ is a stack
(using $\bullet$ for push and $[\,]$ for empty stack)
of frames of unfinished function calls.  A
frame is a triple $(e,x,\rho)$ representing the call site
$
(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
$
being evaluated in environment $\rho$.
Frames can also be viewed as continuations, in this view
the ({\sc ord-return}) 
 rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.

@@

% A program terminates successfully if it reaches the halt state
% in Figure~\ref{fig:lang-opsem-small}.
% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
% is the halt state (with result value $\rho(x)$).

\begin{figure}[t!]  
% (Random text to ensure proper line-spacing (problem with infrule macro?).)

\footnotesize
\begin{minipage}{0.335\textwidth}
\infrule[ord-const]
        {}
        {\rho, \stk, \heap,   \pp{\pi}{k}
          \rightsquigarrow \rho, \stk, \heap, k}
\end{minipage}\begin{minipage}{0.655\textwidth}
\infrule[ord-cons]
      {
%%\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
}
      {\rho, \ell \bullet \stk,\heap, \pp{\pi}{(\CONS\,\, x\ y)}
        \rightsquigarrow \rho, \stk, \heap[\ell\mapsto(\rho(x),\rho(y))],
        \ell}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-car]
        {\heap(\rho(x)) = (e_1, e_2)}
        {\rho, \stk, \heap, \pp{\pi}{(\CAR\,\, x)}
          \rightsquigarrow \rho, \stk, \heap, e_1}
\end{minipage}
\begin{minipage}{0.395\textwidth}
\infrule[ord-car]
        {\heap(\rho(x)) = e}
        {\rho, \stk, \heap, \pp{\pi}{(\CAR\,\, x)}
          \rightsquigarrow \rho, (\CAR \bullet):\stk, \heap, v_1}
\end{minipage}\\[0.5ex]

\begin{minipage}{0.595\textwidth}
\infrule[ord-cdr]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
          \rightsquigarrow \heap, v_2}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-id]
      {
}
      {\rho, \heap, \pp{\pi}{(\ID\,\, x)}
        \rightsquigarrow \heap, \rho(x)}
\end{minipage}\begin{minipage}{0.595\textwidth}
\infrule[ord-prim]
      {\rho(x) \in \mathbb{N} \andalso \rho(y)\in \mathbb{N}}
      {\rho, \heap, \pp{\pi}{(\PRIM\,\, x\ y)}
        \rightsquigarrow \heap, \rho(x) + \rho(y)}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
% \infrule[ord-null-false]
\infrule[]
        {\rho(x)\not= \NIL
        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
          \rightsquigarrow \heap, 0}
\end{minipage}\begin{minipage}{0.595\textwidth}
% \infrule[ord-null-true]
\infrule[ord-null]
        {\rho(x)= \NIL
        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
          \rightsquigarrow \heap, 1}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
% \infrule[ord-if-true]{
\infrule[]{
  \rho(x) \in \mathbb{N}\setminus\{0\}
}{
  \rho,\stk, \heap,
  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
  \longrightarrow \rho,\stk, \heap, e_1}
\end{minipage}\begin{minipage}{0.595\textwidth}
% \infrule[ord-if-false]{
\infrule[ord-if]{
  \rho(x) = 0
}{
  \rho,\stk, \heap,
  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
  \longrightarrow \rho,\stk, \heap, e_2}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-let-nonfn]{
%  \rho,\stk, \heap, \pp{\pi_1}{s} \longrightarrow \rho,\stk, \heap', v_1 
  \rho, \heap, \pp{\pi_1}{s} \rightsquigarrow \heap', v 
  \andalso
  \mbox{$s$ is not
$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
       }
}{
  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
    \pp{\pi_2}{e})}
  \longrightarrow
  \rho[x \mapsto v], \stk, \heap', \pp{\pi_2}{e}}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-let-fncall]{
  \mbox{$s$ is 
$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
       }
\andalso
\mbox{$f$ defined as $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \pp{\pi_f}{e_f})$}
}{
  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
    \pp{\pi_2}{e})}
  \longrightarrow
% [z_1\mapsto \rho(y_1), \ldots, z_n\mapsto \rho(y_n)], 
[\myvec{z} \mapsto \rho(\myvec{y})],
  \,(\rho,x,e)\bullet\stk,\, \heap, e_f
}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-return]{
}{
  \rho, \,(\rho',x',e')\bullet\stk,\, \heap,
  \pp{\pi}{(\RETURN\ x)}
  \longrightarrow \rho'[x' \mapsto \rho(x)], \stk, \heap, e'}
\end{minipage}

% \normalsize
% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
% is the halt state (with result value $\rho(x)$).

  \caption{The small-step operational semantics}
  \label{fig:lang-opsem-small}
\figrule
\normalsize
\end{figure}

%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point

>>>>

\subsubsection{Semantics}
We now give an operational
semantics for  our language.  Later, a refinement of
the  operational semantics,  which we  call minefield  semantics, will
serve to prove liveness analysis correct.
We give a small-step semantics because, unlike big-step semantics,
correctness for non-terminating programs does not need special treatment.
We start with the domains used by the semantics:
\[
\begin{array}{rlcl@{\hspace{2em}}l}
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Loc} & \mbox{-- Environment} \\ 
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Val \times
  Val}& \mbox{-- Values}\\
c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (\mathit{Val} +
\mathit{Clo}) & \mbox{-- Heap}
\end{array}
\]
%Note that in the eager case, if a cons had a numeric argument, this
%fact was known when the cons was created. Thus you could have unboxed
%numbers. This is not possible now. For instance, the argument of a
%cons could be a closure which evaluates to a number.
%Decision - numbers are boxed
Here $\mathit{Loc}$ is countable set  of locations in the heap.  Since
all data objects are boxed, we  model an environment as a mapping from
the set of variables of the program $\mathit{Var}$ to locations in the
heap.  These either contain a value  in WHNF (a number, the empty list
$\NIL$, or  a \CONS\ cell)  or a {\em  closure}.  A closure is  a pair
$(s, \rho)$ in which $s$ is an unevaluated application with the values
of its free variables supplied by $\rho$.

Finally, $\stk$  is a stack (using  $\bullet$ for push  and $[\,]$ for
empty stack)  of frames  of unfinished function  calls.  A frame  is a
triple  $(e,x,\rho)$   representing  the  call  site   $  (\LET\,\,  x
\leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\,  e) $ being evaluated in
environment $\rho$.   Frames can also  be viewed as  continuations, in
this view  the ({\sc ord-return})  rule in the  small-step operational
semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.

@@

\begin{figure}[t!]
\begin{tabular}{|l|l|c|}
\hline
Condition & Transition & Rule name \\
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) \mapsto \kappa], e$    &  \sc{const} \\
\hline
          & \shortstack[l]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
$ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (v_1, v_2) & \rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v_1], e$      &  \sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (e, \rho') & \rho, S, H, (\CAR~x)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \shortstack[l]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (e, \rho') & \rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } (e, \rho') & \rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
\hline
\shortstack[l]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
[\rho(\myvec{x})/\myvec{y}], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
\hline
$\ell$ is a new location& \shortstack[l]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$ \\
$[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
  \ell)\oplus\lfloor \rho\rfloor_{FV(S)-\{x\}})], e$} & \sc{let} \\ 
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\ 
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\ 
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[l]{$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
$\rho'\oplus\rho, (\rho, x, (\IF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
\hline
\shortstack[l]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
(\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[l]{$\rho, S, H, (\RETURN~x)
  \rightsquigarrow$ \\
$\rho'\oplus\rho,~ (\rho, x, (\RETURN~x))\!:\!S, H,  e$} & \sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language}\label{fig:lang-semantics}
\end{figure}
%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
